---
title: "Analyzing World Happiness"
author: "Ingrid Zoll, Christy Lei & Nat Rubin"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: hpstr
    highlight: github
    
---
(feel free to edit the paragraphs here!)

This year the **International Day of Happiness** was celebrated on March 20 with the theme "Happiness For All, Together” while (ironically) the world is coping with a stressful pandemic. But don't let that get on your nerves too much - we are about look at some happiness-related data to (hopefully) get some clues as to why some countries are happier than others.

Specifically, we want to analyze and visualize the relationships between **happiness score**, **suicide rates**, and **mental health access** for all the countries in the recent years (2015-2019).

## Data Wrangling

```{r setup, warning=FALSE, message=FALSE}
library(knitr)
library(tidyverse)
library(ggplot2)
library(viridis)
library(ggthemes)
library(plotly)
library(countrycode)
library(gridExtra)

```

Describe our data sources: e.g. our five happiness datasets (2015-2019) are obtained from Kaggle (https://www.kaggle.com/unsdsn/world-happiness)...

```{r warning=FALSE, message=FALSE}
happiness_2015 <- read.csv("https://raw.githubusercontent.com/Reed-Statistics/math241S20PostGrp4/master/2015.csv?token=ANDSWNJBOL4TOHNDMTCN2B26PZABE")
happiness_2016 <- read.csv("https://raw.githubusercontent.com/Reed-Statistics/math241S20PostGrp4/master/2016.csv?token=ANDSWNPJDSBHOCDJ3XFEZGK6PZADI")
happiness_2017 <- read.csv("https://raw.githubusercontent.com/Reed-Statistics/math241S20PostGrp4/master/2017.csv?token=ANDSWNLSUG5QK23B6O6CWXS6PZAFA")
happiness_2018 <- read.csv("https://raw.githubusercontent.com/Reed-Statistics/math241S20PostGrp4/master/2018.csv?token=ANDSWNJIBN65O6CBRQRR2726PZAFW")
happiness_2019 <- read.csv("https://raw.githubusercontent.com/Reed-Statistics/math241S20PostGrp4/master/2019.csv?token=ANDSWNO6U5A4PHJ3QJBMZTS6PZAGM")
mental_health_facilities <- read.csv("https://raw.githubusercontent.com/Reed-Statistics/math241S20PostGrp4/master/mental_health_facilities.csv?token=ANDSWNOJDJM2QZZHRZY6XRC6PZAIA")
suicide_death_rates <- read.csv("https://raw.githubusercontent.com/Reed-Statistics/math241S20PostGrp4/master/suicide-death-rates.csv?token=ANDSWNKAB4OYFH5KITURBOK6PZAJG")
#time_series_deaths <- read.csv("")
```

Walk the readers through the data wrangling processes (e.g. select relevant columns/rows, rename variables, etc.)
```{r warning=FALSE, message=FALSE}

lower_bound <- rep(NA, 158)
upper_bound <- rep(NA, 158)

happiness_2015 <- happiness_2015 %>%
  mutate(year = 2015) %>%
  select(1, 3, 4, 6:11, 13) %>%
  mutate(lower_bound = lower_bound, upper_bound = upper_bound) %>%
  rename(country = 1,
         happiness_rank = 2,
         happiness_score = 3,
         gdp = 4,
         family = 5,
         life_expectancy = 6,
         freedom = 7,
         trust_corruption = 8,
         generosity = 9,
         year = 10,
         lower_whisker = 11,
         upper_whisker = 12)

happiness_2016 <- happiness_2016 %>%
  mutate(year = 2016) %>%
  select(1, 3, 4:12, 14) %>%
  rename(country = 1,
         happiness_rank = 2,
         happiness_score = 3,
         gdp = 6,
         family = 7,
         life_expectancy = 8,
         freedom = 9,
         trust_corruption = 10,
         generosity = 11,
         year = 12,
         lower_whisker = 4,
         upper_whisker = 5)
  

happiness_2017 <- happiness_2017 %>%
  mutate(year = 2017) %>%
  select(-12) %>%
  rename(country = 1,
         happiness_rank = 2,
         happiness_score = 3,
         gdp = 6,
         family = 7,
         life_expectancy = 8,
         freedom = 9,
         trust_corruption = 10,
         generosity = 11,
         year = 12,
         lower_whisker = 4,
         upper_whisker = 5)

lower_bound <- rep(NA, 156)
upper_bound <- rep(NA, 156)

happiness_2018 <- happiness_2018 %>%
  mutate(year = 2018) %>%
  mutate(lower_bound = lower_bound, upper_bound = upper_bound) %>%
  rename(country = 2,
         happiness_rank = 1,
         happiness_score = 3,
         gdp = 4,
         family = 5,
         life_expectancy = 6,
         freedom = 7,
         trust_corruption = 9,
         generosity = 8,
         year = 10,
         lower_whisker = 11,
         upper_whisker = 12)

happiness_2019 <- happiness_2019 %>%
  mutate(year = 2019)  %>%
  mutate(lower_bound = lower_bound, upper_bound = upper_bound) %>%
  rename(country = 2,
         happiness_rank = 1,
         happiness_score = 3,
         gdp = 4,
         family = 5,
         life_expectancy = 6,
         freedom = 7,
         trust_corruption = 9,
         generosity = 8,
         year = 10,
         lower_whisker = 11,
         upper_whisker = 12)
  
happiness <- do.call("rbind", list(happiness_2015, happiness_2016, happiness_2017, happiness_2018, happiness_2019))
```

```{r}
suicide_death_rates <- suicide_death_rates %>%
  select(-2) %>%
  rename(sdr = 3)

countries <- happiness %>%
  left_join(mental_health_facilities, by = c("country" = "Country")) %>%
  left_join(suicide_death_rates, by = c(c("country" = "Entity"),
                                        c("year" = "Year"))) %>%
  select(-13)
  
```

## A first look at the World Happiness Ranking

Write a paragraph to explain the processes of averaging the happiness scores across the years and briefly discuss the most happy vs the least happy countries (maybe map another variable too?). 
```{r fig.width= 12, fig.height = 30}
avg_happiness <- countries %>%
  group_by(country) %>%
  summarize(avg_happiness = mean(happiness_score)) %>%
  as.data.frame()

#make diverging barcharts
avg_happiness$happiness_z <- round((avg_happiness$avg_happiness- 
                                mean(avg_happiness$avg_happiness))/sd(avg_happiness$avg_happiness), 2)
avg_happiness$happiness_type <- ifelse(avg_happiness$happiness_z < 0, "below", "above")
avg_happiness <- avg_happiness[order(avg_happiness$happiness_z), ]
avg_happiness$country <- factor(avg_happiness$country, levels = unique(avg_happiness$country))

happy_plot <- ggplot(avg_happiness, aes(x = country, y = happiness_z, label = happiness_z)) +
         geom_bar(stat = "identity", aes(fill = happiness_type)) +
         scale_fill_manual(name="Happiness Score", 
                    labels = c("Above Average", "Below Average"), 
                    values = c("above"="#DB7093", "below"="#D8BFD8")) + 
         coord_flip() +
         theme_bw() +
         theme_fivethirtyeight() + 
         labs(title = "Average of World Happiness 2015-2019")

happy_plot
```

Write a paragraph to describe the change in suicide rate/happiness by continent.

```{r warning=FALSE, message=FALSE}
countries$continent <- countrycode(sourcevar = countries[, "country"],
                            origin = "country.name",
                            destination = "continent") 

# manually assign the continent name to Kosovo
countries[countries$country =="Kosovo", "continent"] <- "Europe"

# create a scatterplot to visualize the relationship between happiness and suicide
suicide_happiness <- countries %>% 
  filter(year %in% c(2015, 2016, 2017)) %>%
  ggplot(aes(x = happiness_score, y = sdr, 
            color = continent, frame = year, ids = country)) +
  geom_point(aes(size = life_expectancy), alpha = 0.7) +
  scale_x_log10() +
  labs(x = "Happiness Score", y = "Suicide death rate per 100,000",
       size = NULL, color = "Continent") +
  theme_minimal() +
  scale_color_manual(values = c("#DEB887", "#CD5C5C", "#FFD700", "#ADD8E6", "#9ACD32")) 

ggplotly(suicide_happiness)
```



```{r warning=FALSE, message=FALSE}
# create a scatterplot that shows the relationship between mental health facilities and happiness scores
countries %>% ggplot(aes(x = happiness_score, y = Mental.hospitals..per.100.000.population.)) +
  scale_y_log10() + 
  scale_x_log10() + 
  geom_jitter(alpha = 0.5, color = "#BC8F8F")
```
(we don't have to use this graph - I'm just curious about the log transformed relationship)

=======
Write a paragraph to describe the relationship between mental health access and happiness/suicide (do we need to log transform the axis?)
```{r warning=FALSE, message=FALSE}
#side-by-side comparison of mental health access versus happiness/suicide rate

happiness_vs_health <- countries %>%
  filter(year == 2016) %>%
  ggplot(aes(x = Mental.health.units.in.general.hospitals..per.100.000.population., y = happiness_score), na.rm = TRUE) +
  geom_point(color = "darkorchid2") +
  geom_smooth(method=lm, color = "darkorchid4") +
  theme_bw() +
  labs(x = "Mental Health Units in General Hospitals per 100,000",
       y = "Happiness Score (2016)",
       title = "Mental Health access versus Happiness Score")

suicide_vs_health <- countries %>%
  filter(year == 2016) %>%
  ggplot(aes(x = Mental.health.units.in.general.hospitals..per.100.000.population., y = sdr), na.rm = TRUE) +
  geom_point(color = "royalblue2") +
  geom_smooth(method=lm, color = "royalblue4") +
  theme_bw() +
  labs(x = "Mental Health Units in General Hospitals per 100,000",
       y = "Suicide Death Rate (2016)",
       title = "Mental Health access versus Suicide Death Rate")

happiness_vs_health
suicide_vs_health

```


We also need a table somewhere in our blogpost.
