---
title: "Analyzing World Happiness"
author: "Ingrid Zoll, Christy Lei & Nat Rubin"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: hpstr
    highlight: github
    
---
(feel free to edit the paragraphs here!)

This year the **International Day of Happiness** was celebrated on March 20 with the theme "Happiness For All, Together” while (ironically) the world is coping with a stressful pandemic. But don't let that get on your nerves too much - we are about look at some happiness-related data to (hopefully) get some clues as to why some countries are happier than others.

Specifically, we want to analyze and visualize the relationships between **happiness score**, **suicide rates**, and **mental health access** for all the countries in the recent years (2015-2019).

## Data Wrangling

```{r setup, warning=FALSE, message=FALSE}
library(knitr)
library(tidyverse)
library(ggplot2)
library(viridis)
library(ggthemes)
library(plotly)
library(countrycode)
library(gridExtra)

```

Describe our data sources: e.g. our five happiness datasets (2015-2019) are obtained from Kaggle (https://www.kaggle.com/unsdsn/world-happiness)...

```{r warning=FALSE, message=FALSE}
happiness_2015 <- read.csv("https://raw.githubusercontent.com/Reed-Statistics/math241S20PostGrp4/master/2015.csv?token=ANDSWNJBOL4TOHNDMTCN2B26PZABE")
happiness_2016 <- read.csv("https://raw.githubusercontent.com/Reed-Statistics/math241S20PostGrp4/master/2016.csv?token=ANDSWNPJDSBHOCDJ3XFEZGK6PZADI")
happiness_2017 <- read.csv("https://raw.githubusercontent.com/Reed-Statistics/math241S20PostGrp4/master/2017.csv?token=ANDSWNLSUG5QK23B6O6CWXS6PZAFA")
happiness_2018 <- read.csv("https://raw.githubusercontent.com/Reed-Statistics/math241S20PostGrp4/master/2018.csv?token=ANDSWNJIBN65O6CBRQRR2726PZAFW")
happiness_2019 <- read.csv("https://raw.githubusercontent.com/Reed-Statistics/math241S20PostGrp4/master/2019.csv?token=ANDSWNO6U5A4PHJ3QJBMZTS6PZAGM")
mental_health_facilities <- read.csv("https://raw.githubusercontent.com/Reed-Statistics/math241S20PostGrp4/master/mental_health_facilities.csv?token=ANDSWNOJDJM2QZZHRZY6XRC6PZAIA")
suicide_death_rates <- read.csv("https://raw.githubusercontent.com/Reed-Statistics/math241S20PostGrp4/master/suicide-death-rates.csv?token=ANDSWNKAB4OYFH5KITURBOK6PZAJG")
#time_series_deaths <- read.csv("")
```

First we needed to standardize the data across each year. We removed some columns, like `region` and the data authors' `dystopia` measure, which were unhelpful for our analysis. Some years included error bounds on the country's overall happiness score, while others didn't. Furthermore, the names and orders of the variables changed from year to year. We inserted NAs where it was necessary and made sure each year had the same variables with easily workable names before combining them into one dataset.

```{r warning=FALSE, message=FALSE}

lower_bound <- rep(NA, 158)
upper_bound <- rep(NA, 158)

happiness_2015 <- happiness_2015 %>%
  mutate(year = 2015) %>%
  select(1, 3, 4, 6:11, 13) %>%
  mutate(lower_bound = lower_bound, upper_bound = upper_bound) %>%
  rename(country = 1,
         happiness_rank = 2,
         happiness_score = 3,
         gdp = 4,
         family = 5,
         life_expectancy = 6,
         freedom = 7,
         trust_corruption = 8,
         generosity = 9,
         year = 10,
         lower_whisker = 11,
         upper_whisker = 12)

happiness_2016 <- happiness_2016 %>%
  mutate(year = 2016) %>%
  select(1, 3, 4:12, 14) %>%
  rename(country = 1,
         happiness_rank = 2,
         happiness_score = 3,
         gdp = 6,
         family = 7,
         life_expectancy = 8,
         freedom = 9,
         trust_corruption = 10,
         generosity = 11,
         year = 12,
         lower_whisker = 4,
         upper_whisker = 5)
  

happiness_2017 <- happiness_2017 %>%
  mutate(year = 2017) %>%
  select(-12) %>%
  rename(country = 1,
         happiness_rank = 2,
         happiness_score = 3,
         gdp = 6,
         family = 7,
         life_expectancy = 8,
         freedom = 9,
         trust_corruption = 10,
         generosity = 11,
         year = 12,
         lower_whisker = 4,
         upper_whisker = 5)

lower_bound <- rep(NA, 156)
upper_bound <- rep(NA, 156)

happiness_2018 <- happiness_2018 %>%
  mutate(year = 2018) %>%
  mutate(lower_bound = lower_bound, upper_bound = upper_bound) %>%
  rename(country = 2,
         happiness_rank = 1,
         happiness_score = 3,
         gdp = 4,
         family = 5,
         life_expectancy = 6,
         freedom = 7,
         trust_corruption = 9,
         generosity = 8,
         year = 10,
         lower_whisker = 11,
         upper_whisker = 12)

happiness_2019 <- happiness_2019 %>%
  mutate(year = 2019)  %>%
  mutate(lower_bound = lower_bound, upper_bound = upper_bound) %>%
  rename(country = 2,
         happiness_rank = 1,
         happiness_score = 3,
         gdp = 4,
         family = 5,
         life_expectancy = 6,
         freedom = 7,
         trust_corruption = 9,
         generosity = 8,
         year = 10,
         lower_whisker = 11,
         upper_whisker = 12)
  
happiness <- do.call("rbind", list(happiness_2015, happiness_2016, happiness_2017, happiness_2018, happiness_2019))
```

We then added suicide rates and mental health access to the dataset. Unfortunately, mental health access was only collected in one year, so that value remains constant across the years in each country.

```{r warning=FALSE}
suicide_death_rates <- suicide_death_rates %>%
  select(-2) %>%
  rename(sdr = 3)

countries <- happiness %>%
  left_join(mental_health_facilities, by = c("country" = "Country")) %>%
  left_join(suicide_death_rates, by = c(c("country" = "Entity"),
                                        c("year" = "Year"))) %>%
  select(-13)
  
```

So, how happy are countries overall? The World Happiness Report calculates each country's happiness score by using survey data. Citizens of each country are asked where their own life ranks on a scale of 1 to 10, with 10 being the best life possible and 0 being the worst life possible. Their responses are weighted to generate a national score (See Table 1). Overall, people view their lives as being of average to slightly-above average quality. The wide range between the maximum and minimum are telling: people in the happiest countries are much, much happier than those in the least happy countries.

```{r}
happiness_summary <- countries %>%
  summarize(mean(happiness_score), median(happiness_score), sd(happiness_score), min(happiness_score),  max(happiness_score))
options(knitr.table.format = "html")
kable(happiness_summary, digits = 2, 
      col.names = c("Mean", "Median", 
                    "Standard Dev.", "Minimum", "Maximum"),
      caption = "Table 1. Summary Statistics of National Happiness Scores")
```
## A First Look at the World Happiness Ranking

  We want to directly visualize the world happiness ranking using a **diverging bar chart**, which is great for illustrating the spread of negative and positive values. 
  
  First, we created a new dataframe `avg_happiness` that contains the average happiness score for every country from 2015 to 2019.
```{r}
avg_happiness <- countries %>%
  group_by(country) %>%
  summarize(avg_happiness = mean(happiness_score)) %>%
  as.data.frame()
```

  Next, we created a new column `happiness_z` that contains the normalized happiness score ($z = (x-\mu)/ \sigma$) for each country, and another column `happiness_type` to indicate which country is above or below 0 using `ifelse()`.

```{r}
avg_happiness$happiness_z <- round((avg_happiness$avg_happiness- 
                                mean(avg_happiness$avg_happiness))/sd(avg_happiness$avg_happiness), 2)
avg_happiness$happiness_type <- ifelse(avg_happiness$happiness_z < 0, "below", "above")
```

We also sorted the columns by the happiness score and converted the column `country` to factor to retain sorted order in plot.
```{r}
avg_happiness <- avg_happiness[order(avg_happiness$happiness_z), ]
avg_happiness$country <- factor(avg_happiness$country, levels = unique(avg_happiness$country))
```

Now, let's visualize the diverging bars using `geom_bar()`! 
```{r fig.width = 10, fig.height = 25}
happy_plot <- ggplot(avg_happiness, aes(x = country, y = happiness_z, label = happiness_z)) +
         geom_bar(stat = "identity", aes(fill = happiness_type)) +
         scale_fill_manual(name="Happiness Score", 
                    labels = c("Above Average", "Below Average"), 
                    values = c("above"="#DB7093", "below"="#D8BFD8")) + 
         coord_flip() +
         theme_fivethirtyeight() + 
         labs(title = "Average of World Happiness 2015-2019")

happy_plot
```

## Change in world happiness and suicide rate from 2015 to 2017

Then, we wanted to create a **scatterplot** to visualize (1) the change in world happiness and suicide rate from 2015 to 2017 by continent (we don't have data for suicide rate after 2017) and (2) the relationship between happiness rating and suicide rate.

First, we used the library `countrycode` to create a new variable with the name of the continent to which each country belongs.
```{r warning=FALSE, message=FALSE}
countries$continent <- countrycode(sourcevar = countries[, "country"],
                            origin = "country.name",
                            destination = "continent") 
```

```{r echo = FALSE}
# manually assign the continent name to Kosovo
countries[countries$country =="Kosovo", "continent"] <- "Europe"
```

Then, we used `filter()` to only include data from 2015, 2016, and 2017, and visualized the relationship between happiness and suicide using `geom_point()`. We also made use of the function `ggplotly()` from the interative graphing library `plotly` to add a slider that illustrates the change over the years.
```{r}
# create a scatterplot to visualize the relationship between happiness and suicide
suicide_happiness_plot <- countries %>% 
  filter(year %in% c(2015, 2016, 2017)) %>%
  ggplot(aes(x = happiness_score, y = sdr, 
            color = continent, frame = year, ids = country)) +
  geom_point(aes(size = 1, alpha = 0.7)) +
  scale_x_log10() +
  labs(x = "Happiness Score", y = "Suicide death rate per 100,000",
       size = NULL, color = "Continent") +
  theme_minimal() +
  scale_color_manual(values = c("#DEB887", "#CD5C5C", "#FFD700", "#ADD8E6", "#9ACD32")) 

ggplotly(suicide_happiness_plot)
```



```{r warning=FALSE, message=FALSE}
# create a scatterplot that shows the relationship between mental health facilities and happiness scores
# countries %>% ggplot(aes(x = happiness_score, y = Mental.hospitals..per.100.000.population.)) +
 # scale_y_log10() + 
 # scale_x_log10() + 
 # geom_jitter(alpha = 0.5, color = "#BC8F8F")
```
(we don't have to use this graph - I'm just curious about the log transformed relationship)


## Mental health access, happiness, and suicide

How does mental health access factor into happiness and suicide? We used the indiciator of mental health access with the fewest NAs, which is the number of mental health units in general hospitals per 100,000. Because this variable is highly right-skewed, we log it to better visualize the relationship.

The two plots reveal that mental health access appears to have a stronger relationship with happiness than with the suicide death rate. The error bar on the regression line for the suicide rate shows that the trend could be flat or even negative, rather than the small positive relationship seen here. The stronger relationship between mental health access and happiness is unsurprising. Greater access to mental healthcare may increase happiness, or another factor, such as a country's wealth, may connect the two.

```{r warning=FALSE, message=FALSE}
#side-by-side comparison of mental health access versus happiness/suicide rate

happiness_vs_health <- countries %>%
  filter(year == 2016) %>%
  ggplot(aes(x = Mental.health.units.in.general.hospitals..per.100.000.population., y = happiness_score), na.rm = TRUE) +
  geom_point(color = "darkorchid2") +
  geom_smooth(method=lm, color = "darkorchid4") +
  scale_x_log10() +
  theme_bw() +
  labs(x = "(Log of) Mental Health Units in General Hospitals per 100,000",
       y = "Happiness Score (2016)",
       title = "Log of Mental Health access versus Happiness Score")

suicide_vs_health <- countries %>%
  filter(year == 2016) %>%
  ggplot(aes(x = Mental.health.units.in.general.hospitals..per.100.000.population., y = sdr), na.rm = TRUE) +
  geom_point(color = "royalblue2") +
  geom_smooth(method=lm, color = "royalblue4") +
  scale_x_log10() +
  theme_bw() +
  labs(x = "(Log of) Mental Health Units in General Hospitals per 100,000",
       y = "Suicide Death Rate (2016)",
       title = "Log of Mental Health access versus Suicide Death Rate")

happiness_vs_health
suicide_vs_health

```


##What Contributes to Happiness?

##Contributions to happiness

What factors contribute to happiness? We wanted to explore the different contributors to the happiness score and how they vary between the five most happy countries and the five least happy countries. For this, we're only looking at the data from 2019. First, we used `filter()` to select the five happiest and five least happy countries that we wanted to look at. Then we pivoted the 2019 dataset so the individual measurements that contribute to the overall happiness measurement (**family, freedom, economy, generosity, health,** and **trust**) would be observations and not variables. Finally, we made country name a factor so when we graph this data, the countries are still ordered by happiness rank. 

This graph, which is faceted by measurement type and color coded by country, examines the extent to which the following factors contribute to the overall happiness score for a country: **Social support**, **Freedom to make life choices**, **GDP per capita**, **Generosity**, **Healthy life expectancy at birth**, and **Perception of corruption**. This chart shows that **social support**, **GDP**, and **healthy life expentancy** seem to be the largest contributors to how happy a country is. Those three categories show the greatest difference between the happiest and least happy countries, whereas there is much less of a difference in **freedom to make life choices**, **generosity**, and **perception of corruption**. This data indicates that happy countries are more socially supportive, have a higher GDP, and have healthier populations than less happy countries. Since GDP and life expectancy are fact based measurements rather than how the residents of the country feel, I think that social support is the most important factor contributing to the happiness of countries.

More information on each of the factors that contribute to the happiness measurement can be found on the World Happiness Report (https://worldhappiness.report/ed/2019/changing-world-happiness/#technical-box-detailed-information-about-each-of-the-predictors-in-table).


```{r}
top_bottom_2019_happiness<-happiness_2019 %>%
  filter(happiness_rank <= 5 | happiness_rank >= 152) %>% 
  pivot_longer(cols=c(gdp, family, life_expectancy, freedom, generosity, trust_corruption), names_to = "category", values_to = "values") %>% 
  mutate(country = factor(country, levels=c("Finland", "Denmark", "Norway", "Iceland", "Netherlands", "Rwanda", "Tanzania", "Afghanistan", "Central African Republic", "South Sudan")))

ggplot(top_bottom_2019_happiness, mapping=aes(x=category, y=values, fill=country))+
  geom_col(position="dodge", color="black", size=0.25)+
  theme_bw()+
  scale_fill_discrete(name="Country (with rank)", labels = c("Finland (1)", "Denmark (2)", "Norway (3)", "Iceland (4)", "Netherlands (5)", "Rwanda (152)", "Tanzania (153)", "Afghanistan (154)", "Central African\n Republic (155)", "South Sudan (156)"))+
  scale_x_discrete(labels = c("Social\n support", "Freedom to\n make life\n choices", "GDP\n per capita" , "Generosity", "Healthy life\n expectancy\n at birth", "Perception\n of corruption"))+
  labs(x="Measurement Category", y="Importance score")
```
We also need a table somewhere in our blogpost.
